---
title: 'Statistical Rethinking: A Bayesian Course with Examples in R and Stan'
subtitle: 'Chapter 4: Geocentric models'
author: "Marwin Carmo"
date: "12/01/2022"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r packages, include=FALSE}
library(rethinking)
library(ggplot2)
library(dplyr)
```

## 4.1.1. Normal by addition

```{r 4.1}
pos <- replicate(1000, sum(runif(16,-1,1)))

with(density(pos), data.frame(x, y)) |> 
  ggplot(mapping = aes(x = x, y = y)) +
    geom_line(colour = "darkblue", size=1.2) +
    theme_bw(12) +
    labs(y = "Density", x = "position", title = "16 steps")

```

## 4.1.2. Normal by multiplication

```{r 4.3}
growth <- replicate( 10000 , prod( 1 + runif(12,0,0.1) ) )
dens( growth , norm.comp=TRUE )
```
## 4.1.3. Normal by log-multiplication

Adding logs is equivalent to multiplying the original numbers.

```{r 4.5}
log.big <- replicate( 10000 , log(prod(1 + runif(12,0,0.5))) )
dens(log.big, norm.comp = TRUE)
```

## 4.3. Gaussian model of height

```{r 4.7}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
```

Average height prior

```{r 4.12}
curve(dnorm(x, 178, 20), from=100, to = 250, 
      xlab = expression(mu), main = expression(paste(mu, " ~ Normal(178, 20)")))
```

Flat prior for $\sigma$

```{r 4.13}
curve(dunif(x, 0, 50), from=-10, to=60,
      xlab = expression(sigma), main = expression(paste(sigma, " ~ Uniform(0, 50)")))
```

Simulating heights by sampling from the prior

```{r 4.14}
sample_mu <- rnorm( 1e4 , 178 , 20 )
sample_sigma <- runif( 1e4 , 0 , 50 )
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma )
dens( prior_h, xlab = "height", 
      main = expression(paste(h, " ~ Normal(", mu, ",", sigma, ")")))
```

## 4.3.3. Grid approximation of the posterior distribution

```{r 4.16}
mu.list <- seq( from=150, to=160 , length.out=100 )
sigma.list <- seq( from=7 , to=9 , length.out=100 )
post <- expand.grid( mu=mu.list , sigma=sigma.list )
post$LL <- sapply( 1:nrow(post) , function(i) sum(dnorm( d2$height , post$mu[i], 
                                                         post$sigma[i] , log=TRUE ) ) )
post$prod <- post$LL + dnorm( post$mu , 178 , 20 , TRUE ) +
dunif( post$sigma , 0 , 50 , TRUE )
post$prob <- exp( post$prod - max(post$prod) )
```

```{r}
contour_xyz( post$mu , post$sigma , post$prob )
image_xyz( post$mu , post$sigma , post$prob )
```

## 4.3.4. Sampling from the posterior

```{r 4.19}
sample.rows <- sample( 1:nrow(post) , size=1e4 , replace=TRUE ,prob=post$prob )
sample.mu <- post$mu[ sample.rows ]
sample.sigma <- post$sigma[ sample.rows ]

plot( sample.mu , sample.sigma , cex=0.5 , pch=16 , col=col.alpha(rangi2,0.1) )
```

```{r}
PI( sample.mu )
PI( sample.sigma )
```

```{r}
den <- density( sample.mu )
plot(den, xlab= "sample.mu", main = NULL)

l <- min(which(den$x >= PI( sample.mu )[[1]]))
h <- max(which(den$x < PI( sample.mu )[[2]]))

polygon(c(den$x[c(l, l:h, h)]),
        c(0, den$y[l:h], 0),
        col = "slateblue1")
```
```{r}
den <- density( sample.sigma )
plot(den, xlab= "sample.sigma", main = NULL)

l <- min(which(den$x >= PI( sample.sigma )[[1]]))
h <- max(which(den$x < PI( sample.sigma )[[2]]))

polygon(c(den$x[c(l, l:h, h)]),
        c(0, den$y[l:h], 0),
        col = "slateblue1")
```

Overthinking: Sample size and the normality of $\sigma$’s posterior

```{r 4.23}
d3 <- sample( d2$height , size=20 )
```

```{r 4.24}
mu.list <- seq( from=150, to=170 , length.out=200 )
sigma.list <- seq( from=4 , to=20 , length.out=200 )
post2 <- expand.grid( mu=mu.list , sigma=sigma.list )
post2$LL <- sapply( 1:nrow(post2) , function(i)
sum( dnorm( d3 , mean=post2$mu[i] , sd=post2$sigma[i] ,
log=TRUE ) ) )
post2$prod <- post2$LL + dnorm( post2$mu , 178 , 20 , TRUE ) +
dunif( post2$sigma , 0 , 50 , TRUE )
post2$prob <- exp( post2$prod - max(post2$prod) )
sample2.rows <- sample( 1:nrow(post2) , size=1e4 , replace=TRUE ,
prob=post2$prob )
sample2.mu <- post2$mu[ sample2.rows ]
sample2.sigma <- post2$sigma[ sample2.rows ]
plot( sample2.mu , sample2.sigma , cex=0.5 ,
col=col.alpha(rangi2,0.1) ,
xlab="mu" , ylab="sigma" , pch=16 )
dens( sample2.sigma , norm.comp=TRUE )
```

This code will also show a normal approximation with the same mean and variance. Now you can see that the posterior for σ is not Gaussian, but rather has a long tail towards higher values.

## 4.3.5. Finding the posterior distribution with `quap`

```{r 4.27}
flist <- alist(height ~ dnorm( mu , sigma ) ,
               mu ~ dnorm( 178 , 20 ) ,
               sigma ~ dunif( 0 , 50 )
)
```

```{r 4.28}
m4.1 <- quap( flist , data=d2 )
precis(m4.1)
```

Sampling vectors of values from a multi-dimensional Gaussian distribution

```{r}
post <- extract.samples( m4.1 , n=1e4 )
head(post)
precis(post)
```

## 4.4. Linear prediction

```{r 4.37}
plot( d2$height ~ d2$weight , main = "Height x Weight")
```

### 4.4.1.3. Priors

```{r 4.38}
set.seed(2971)
N <- 100 # 100 lines
a <- rnorm( N , 178 , 20 )
b <- rnorm( N , 0 , 10 )
```

```{r}
plot( NULL , xlim=range(d2$weight) , ylim=c(-100,400) , xlab="weight" , ylab="height" )
abline( h=0 , lty=2 )
abline( h=272 , lty=1 , lwd=0.5 )
mtext( "b ~ dnorm(0,10)" )
xbar <- mean(d2$weight)
for ( i in 1:N ) curve( a[i] + b[i]*(x - xbar) ,
                        from=min(d2$weight) , to=max(d2$weight) , add=TRUE , 
                        col=col.alpha("black",0.2) )
```

$$
\beta \sim LogNormal(0,1)
$$

```{r}
b <- rlnorm( 1e4 , 0 , 1 )
dens( b , xlim=c(0,5) , adj=0.1 )
```

```{r 4.41}
set.seed(2971)
N <- 100
# 100 lines
a <- rnorm( N , 178 , 20 )
b <- rlnorm( N , 0 , 1 )

plot( NULL , xlim=range(d2$weight) , ylim=c(-100,400) , xlab="weight" , ylab="height" )
abline( h=0 , lty=2 )
abline( h=272 , lty=1 , lwd=0.5 )
mtext( "b ~ dnorm(0,10)" )
xbar <- mean(d2$weight)
for ( i in 1:N ) curve( a[i] + b[i]*(x - xbar) ,
                        from=min(d2$weight) , to=max(d2$weight) , add=TRUE , 
                        col=col.alpha("black",0.2) )
```
## 4.4.2. Finding the posterior distribution

```{r 4.42}
# define the average weight, x-bar
xbar <- mean(d2$weight)

# fit model
m4.3 <- quap(alist(height ~ dnorm( mu , sigma ) ,
                   mu <- a + b*( weight - xbar ) ,
                   a ~ dnorm( 178 , 20 ) ,
                   b ~ dlnorm( 0 , 1 ) ,
                   sigma ~ dunif( 0 , 50 )) ,
             data=d2 )
```

