---
title: "Statistical Rethinking Winter 2022"
subtitle: "Homework, week 1"
author: "Marwin Carmo"
date: "09/01/2022"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 01
Suppose the globe tossing data (Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution, using grid approximation. Use the same flat prior as in the book.

```{r ex01}
# define grid 
p_grid <- seq( from=0 , to=1 , length.out=50 )
# define prior 
prior <- rep( 1 , 50 )
# compute likelihood at each value in grid 
likelihood <- dbinom( 4 , size=15 , prob=p_grid )
# compute product of likelihood and prior 
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1 
posterior <- unstd.posterior / sum(unstd.posterior)
```

```{r posterior plot}
plot( p_grid , posterior , type="b" , xlab="probability of water" , ylab="posterior probability" )
mtext( "50 points" )
```

## 02

Now suppose the data are 4 water and 2 land. Compute the posterior again, but this time use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the Earth’s surface is water.

```{r ex02}
# define grid 
p_grid <- seq( from=0 , to=1 , length.out=50 )
# define prior 
prior <- ifelse(p_grid < 0.5, 0, 1)
# compute likelihood at each value in grid 
likelihood <- dbinom( 4 , size=6 , prob=p_grid )
# compute product of likelihood and prior 
unstd.posterior <- likelihood * prior
# standardize the posterior, so it sums to 1 
posterior <- unstd.posterior / sum(unstd.posterior)

plot( p_grid , posterior , type="b" , xlab="probability of water" , ylab="posterior probability" )
mtext( "50 points" )
```
